---
title: "Regression Modeling with `mtcars` Dataset"
author: "Stephanie Stallworth"
date: "March 31, 2017"
output: html_document
---

```{r setup, include=FALSE, message = FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###**Executive Summary**

This analysis seeks to answer the following questions about the `mtcars` data set:  
1. Is an automatic or manual transmission better for MPG?  
2. Can the MPG difference between automatic and manual transmission be quantified?

**It can be concluded that manual transmissions are in fact better, with an MPG 1.8 greater than automatic transmissions.**  
  My analysis is described below.

###**Data Processing**
```{r, comment = ""}
#Read data
data(mtcars)

#Examine data structure
str(mtcars)

#Coerce appropriate variables into factors
mtcars$cyl<-factor(mtcars$cyl)
mtcars$vs<-factor(mtcars$vs)
mtcars$gear<-factor(mtcars$gear)
mtcars$carb<-factor(mtcars$carb)
mtcars$am<-factor(mtcars$am,labels = c("Auto","Manual"))

#Confirm variable conversion
str(mtcars)
```

###**Exploratory Analysis**

```{r}
library(ggplot2)

ggplot(mtcars,aes(x = am, y = mpg)) +
        geom_boxplot(fill = c("darkgreen","gold"))+
        labs(x = "Transmission Type",y="Miles Per Gallon", title = "Miles Per Gallon by Transmission Type" )
```


**Conclusion 1: As shown by the plot above, Manual transmissions are better for MPG as these cars have a higher median MPG relative to automative transmissions.**  

###**Regression Analysis**

My exploratory analysis indicates there is a difference in MPG between automatic and manual transmissions.  Thus, the next step is to fit a regression model to determine their relationship.  

```{r, comment =""}
aggregate(mpg~am, data = mtcars, mean)
```

Based on the table above, I hypothesize that the MPG for automatic cars is 7.2 MPG lower than manual cars. A t-test was performed to determine the statistical significance of this difference.
```{r, comment = ""}
auto<-mtcars[mtcars$am=="Auto",]
man<-mtcars[mtcars$am=="Manual",]
t.test(auto$mpg, man$mpg)
```

A p value of 0.00137 is less than the .05 cutoff for statistical significance. Thus, we can reject the null hypothesis and conclude that there is a **statistically significant difference** between the average MPG in automatic cars versus manual cars. A linear model was fitted to quantify this difference.

```{r, comment=""}
fit<-lm(mpg~am, mtcars)
summary(fit)
```
The T-test indicates the average MPG for automatic transmission is 17.1 MPG, with manual transmission being 7.2 MPG higher. However, the R-Squared value indicates this model only explains 36% of the variance. A better option would be to build a multivariate linear regression model to quantify the difference.  

Thus, a pairs plot was generated to determine the additional variables to include in the new model. 
```{r}
pairs(mpg ~., data =mtcars)
```

Based on the plot, the following variables have a very strong correlation with `mpg`: `cyl`, `disp`, `hp`, and `wt`. A new model was fitted using these additional variables and results were compared to the initial model using the `Anova` command.  
```{r, comment =""}
fit2<-lm(mpg~am + cyl + disp + hp + wt, data = mtcars)
summary(fit2)
anova(fit,fit2)
```

The second model's p value of 8.637e-08 suggests adding multivariate is significantly better than using the simple model that was fitted initially. Also, the second model has a higher R-Squared value of  0.8664. Thus, 86.6% of the variance between MPG in automatic and manual transmission is explained by the second model, a significant improvement from the first. The second model's superior performance was further confirmed by a plot of its residuals.
```{r}
par(mfrow=c(2,2))
plot(fit2)
```


**Conclusion 2: The difference in MPG between automatic and manual transmissions can be quantified as 1.81MPG**



