---
## **Predictive Text Exploratory Analysis**
Stephanie Stallworth  
May 7, 2017

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment="",warning=FALSE,message=FALSE,cache=TRUE)
```

### **Executive Summary**
This exploratory analysis is the first step in the development of a predictive text model that will suggest words based on user input.  
The model will be trained using a corpus (a collection of English text) compiled from 3 sources - news, blogs, and tweets. In the following report, I will load and clean the data as well as use NLM (Natural Language Processing) techniques to perform exploratory analysis and build the predictive model.  

### **Processing the Data**

The zip file containing the raw corpus data was downloaded from:   
https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip.


```{r}
# Download and unzip file
if (!file.exists("Coursera-SwiftKey.zip")) {
  download.file("https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip")
  unzip("Coursera-SwiftKey.zip")
}
```
The dataset consists of text from 3 different sources: News, Blogs, and Twitter feeds which are stored locally as:

Blog: ./final/en_US.blogs.txt

News: ./final/en_US.news.txt

Twitter: ./final/en_US.twitter.txt
```{r}
# Read the blogs and Twitter data into R
blogs <- readLines("final/en_US/en_US.blogs.txt", encoding = "UTF-8", skipNul = TRUE)
news <- readLines("final/en_US/en_US.news.txt", encoding = "UTF-8", skipNul = TRUE)
twitter <- readLines("final/en_US/en_US.twitter.txt", encoding = "UTF-8", skipNul = TRUE)
```
A basic summary of the Megabytes, number of lines, number of words, and average length of entry for each file is below:      

```{r}
library(stringi)

# Get file sizes
blogs.size <- file.info("final/en_US/en_US.blogs.txt")$size / 1024 ^ 2
news.size <- file.info("final/en_US/en_US.news.txt")$size / 1024 ^ 2
twitter.size <- file.info("final/en_US/en_US.twitter.txt")$size / 1024 ^ 2

# Get words in files
blogs.words <- stri_count_words(blogs)
news.words <- stri_count_words(news)
twitter.words <- stri_count_words(twitter)

# Summary of the data sets
data.frame(source = c("blogs", "news", "twitter"),
           file.size.MB = c(blogs.size, news.size, twitter.size),
           num.lines = c(length(blogs), length(news), length(twitter)),
           num.words = c(sum(blogs.words), sum(news.words), sum(twitter.words)),
           mean.num.words = c(mean(blogs.words), mean(news.words), mean(twitter.words)))
```
### **Cleaning The Data**
Before performing the exploratory analysis, the data must be scrubbed of all URLs, special characters, punctuation, numbers, trailing spaces, and stop words.

As the summary table shows, the data is quite large. Thus, a sample of 1% is randomly chosen for cleaning and exploratory analysis.
```{r}
# Load library
library(tm);library(stringi)

#blogs   <- iconv(blogs,   "UTF-8", "ASCII", "byte")
news    <- iconv(news,    "UTF-8", "ASCII", "byte")
twitter <- iconv(twitter, "UTF-8", "ASCII", "byte")

set.seed(1332)
data.sample <- c(sample(blogs,   length(blogs)   * 0.01),
                 sample(news,    length(news)    * 0.01),
                 sample(twitter, length(twitter) * 0.01))
#data.sample <- sample(c(blogs, news, twitter), 50000 )
rm(blogs)
rm(news)
rm(twitter)

# Create corpus and clean the data
corpus <- VCorpus(VectorSource(data.sample))
corpus <- tm_map(corpus, tolower)
corpus <- tm_map(corpus, removePunctuation)
corpus <- tm_map(corpus, removeNumbers)
corpus <- tm_map(corpus, stripWhitespace)
corpus <- tm_map(corpus, PlainTextDocument)
```


### **Exploratory Analysis**

**Word Cloud**

Once the data is tidyed, a word cloud is created to get a quick look at the most frequently occurring words in the sampled text, excluding stop words (i.e. the, for, that,etc).
```{r}
library(wordcloud)
tcorpus<-tm_map(corpus, removeWords, stopwords("english"))
wordcloud(tcorpus, max.words=100, random.order=FALSE, colors=brewer.pal(8,"Dark2"))
```
**N-Gram Word Frequency**  

To explore the data further, the most commonly used unigrams, bigrams, and trigrams are then identified.
```{r}
# Load libraries
library(RWeka); library(ggplot2)

# List most common unigrams, bigrams, and trigrams
options(mc.cores=1)

unigram <- function(x) NGramTokenizer(x, Weka_control(min = 1, max = 1))
bigram  <- function(x) NGramTokenizer(x, Weka_control(min = 2, max = 2))
trigram <- function(x) NGramTokenizer(x, Weka_control(min = 3, max = 3))

getFreq <- function(tdm) {
  freq <- sort(rowSums(as.matrix(tdm)), decreasing = TRUE)
  return(data.frame(word = names(freq), freq = freq))
}

makePlot <- function(data, label) {
  ggplot(data[1:50,], aes(reorder(word, -freq), freq)) +
         labs(x = label, y = "Frequency") +
         theme(axis.text.x = element_text(angle = 60, size = 12, hjust = 1)) +
         geom_bar(stat = "identity", fill = I("grey50"))
}
```

**N-Gram Frequency Visualization**

The 50 most common unigrams, bigrams, and trigrams are then visualized in histograms.

```{r}
# 50 most common unigrams plot
freq1 <- getFreq(removeSparseTerms(TermDocumentMatrix(corpus, control=list(tokenize = unigram)), 0.99))
makePlot(freq1, "50 Most Common unigrams")
```
```{r}
# 50 most common bigrams plot
freq2 <- getFreq(removeSparseTerms(TermDocumentMatrix(corpus, control = list(tokenize = bigram)), 0.999))
makePlot(freq2, "50 Most Common bigrams")
```

```{r}
# 50 most common trigrams plot
freq2 <- getFreq(removeSparseTerms(TermDocumentMatrix(corpus, control = list(tokenize = trigram)), 0.999))
makePlot(freq2, "50 Most Common bigrams")
```

### **Conclusion and Next Steps**

Through this exploratory analysis, I identified the most commonly used words which will be helpful in the next steps of this project.   

My plan for the predictive algorithm and app is to use the n-gram model with frequency lookup similar to the exploratory analysis above. It will first use the trigram model to make predictions based on user input. If no matching trigrams can be found, the algorithm will roll back to the bigram model and lastly the unigram model if needed.





