---
title: 'Titanic: Machine Learning from Disaster'
author: "Stephanie Stallworth"
date: "April 17, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment ="",message = FALSE, warning = FALSE)
```

### **Executive Summary**

The sinking of the RMS Titanic is one of the most infamous shipwrecks in history.  On April 15, 1912, during her maiden voyage, the Titanic sank after colliding with an iceberg, killing 1502 out of 2224 passengers and crew. This sensational tragedy shocked the international community and led to better safety regulations for ships.

One of the reasons that the shipwreck led to such loss of life was that there were not enough lifeboats for the passengers and crew. Although there was some element of luck involved in surviving the sinking, some groups of people were more likely to survive than others, such as women, children, and the upper-class.

For this challenge, I will analyze what sorts of people were likely to survive and apply tools of machine learning to predict which passengers survived the tragedy.

### **Data Preprocess**

```{r}

# Read training and test data
train<-read.csv("train.csv")
test<-read.csv("test.csv")

# Examine data structure of training set
str(train)

# Convert variables to appropriate type
train$Survived<-as.factor(train$Survived)
train$Pclass<-as.factor(train$Pclass)

# Confirm variables have beenc converted
str(train)

# Summary Statistics
summary(train)


```


### **Decision Tree**

```{r}
# Install and load required packages for fancy decision tree plotting

library(rpart)
library(rattle)
library(rpart.plot)
library(RColorBrewer)

#Decision Tree
fit <- rpart(Survived ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked, data=train, method="class")
# And then make it look better with fancyRpartPlot!
fancyRpartPlot(fit)





```



