---
title: "Activity Tracker Exploratory Analysis"
author: "Stephanie Stallworth"
date: "April 14, 2017"
output: 
  html_document: 
    keep_md: true
---

```{r setup, include=FALSE, comment = ""}
knitr::opts_chunk$set(echo = TRUE)
```

##**Data Processing**
```{r,comment=""}
# Unzip and read file
unzip("activity.zip")
activity<-read.csv("activity.csv")

# Format dates
activity$day <- weekdays(as.Date(activity$date))
activity$DateTime<-as.POSIXct(activity$date, format = "%Y-%m-%d")

# Omit NA
tActivity <- activity[!is.na(activity$steps), ]
```


##**Analysis**

###**What is mean total number of steps taken per day?**

```{r,comment =""}
# Calculate total number of steps taken per day
sumTable <-  aggregate(steps ~ date, tActivity,sum)
colnames(sumTable) <- c("Date","Steps")
sumTable

# Create histogram of total number of steps taken per day
hist(sumTable$Steps, breaks = 5, xlab = "Steps", main = "Total Steps Per Day", col = "blue")

# Calculate and report the mean and median of the total steps taken per day
as.integer(mean(sumTable$Steps))

# Calculate median of the total steps taken per day
as.integer(median(sumTable$Steps))
```
###**What is the average daily activity pattern?**

```{r}
# Time series plot of the 5-minute intervals and the average number of steps taken, averaged across all days

interval <- aggregate(steps ~ interval, tActivity, mean)
plot(interval$interval, 
     interval$steps, 
     type = "l",
     xlab = "Interval",
     ylab = "# of Steps",
     main = "Average Number of Steps Per Day By Interval")
```

###**When missing value are imputed, do mean values differ from prior calculations? What is the impact of imputing missing data on the estimates of the total daily number of steps?**

There are a number of days/intervals with missing values (coded as `NA`). The presence of missing days may introduce bias into summary calculations so a strategy was devised to address this missing data.

```{r, comment =""}
# Calculate total number of missing values in the dataset 
sum(!complete.cases(activity))

# Average number of steps per day and interval
avgTable <- aggregate(steps ~ interval + day, tActivity, mean)

# Create dataset with NAs substituted
naData <- activity[is.na(activity$steps), ]

# Merge NA with average week day interval table to impute
newData <- merge(naData, avgTable, by = c("interval","day"))

```


A new data set that is equal to the original data set was then created the missing data imputed.
```{r,comment=""}
# Fix original data set with missing values imputed

newData2 <- newData[,c(6,4,1,2,5)]
colnames(newData2) <- c("steps","date", "interval","day","DateTime")

# Merge NA averages and non NA data together
mergeData <- rbind(tActivity,newData2)

# Create sum of steps per date to compare with step 1
sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum)
colnames(sumTable2)<- c("Date", "Steps")
```

Summary calculations were then performed on the new data set with no missing values.

```{r, comment=""}
# Mean of number of steps with NA imputed
as.integer(mean(sumTable2$Steps))

# Median of Steps with NA data imputed
as.integer(median(sumTable2$Steps))


# Histogram of the total number of steps taken each day with NA data imputed
hist(sumTable2$Steps, breaks = 10, col = "blue", xlab = "Daily Steps", main = "Total Steps In A Day")
hist(sumTable$Steps, breaks = 10, col = "grey", xlab = "Daily Steps", main = "Total Steps In A Day", add=T) 
legend("topleft", c("Imputed", "NA"), fill = c("blue","grey"))
```

###**Are there differences in activity patterns between weekdays and weekends?**

```{r}
# Create new factor variable with two levels indicating whether a given date is a weekday or weekend day.

mergeData$DayCategory <- ifelse(mergeData$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")

# Create panel plot containing a time series plot of the 5-minute intervals and the average number of steps taken, averaged across all weekday days or weekend days.

library(ggplot2)

intervalTable2 <- aggregate(steps ~ interval + DayCategory,mergeData,mean)
ggplot(intervalTable2,aes(x = interval, y =steps))+geom_line() + facet_grid(DayCategory~.)+
    xlab("5 minute intervals")+
    ylab("Number of Steps")

```



