---
title: "Motor Trend Cars Presentation & Plotly"
author: "Stephanie Stallworth"
date: "April 13, 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

This is an R Markdown presentation created using **Knit**, which allows the capability to include content as well as output of embedded R code chunks within a document.  

In this presentation, I will also demonstrate how a `ggplot` object can be turned into an interactive web graphic using the `plotly` package.  

##Plotly Code
```{r, eval=FALSE, echo = TRUE}

#Load packages
library(plotly);library(dplyr);library("datasets")

#Read dataset
data(mtcars)

#Create plot
p <-ggplot(mtcars, aes(x = mpg, y = wt, col = factor(cyl))) + 
        geom_point() + 
        labs(title = "Weight vs Mpg", x = "Mpg",y= "Weight") +
        theme(legend.title = element_blank())

#Pass ggplot object to plotly function and format
ggplotly(p) %>%  add_annotations( text="Cylinder", 
                  xref="paper", yref="paper",
                  x=1.02, xanchor="left",
                  y=0.8, yanchor="bottom",   
                  legendtitle=TRUE, showarrow=FALSE ) %>%
                  layout( legend=list(y=0.8, yanchor="top" ) )
```




##mtcars Interactive Plotly Scatter
```{r message = FALSE, warning=FALSE}
library(plotly);library("datasets");data(mtcars);library(dplyr)
p <-ggplot(mtcars, aes(x = mpg, y = wt, col = factor(cyl))) + geom_point() + 
        labs(title = "Weight vs Mpg", x = "Mpg",y= "Weight") +theme(legend.title = element_blank())
ggplotly(p) %>%  add_annotations( text="Cylinder", xref="paper", yref="paper",
                  x=1.02, xanchor="left",
                  y=0.8, yanchor="bottom",    # Same y as legend below
                  legendtitle=TRUE, showarrow=FALSE ) %>%
  layout( legend=list(y=0.8, yanchor="top" ) )
```

***








###Thank You!

