---
title: "Tooth Growth Inferential Analysis"
author: "Stephanie Stallworth"
date: "March 27, 2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Executive Summary**

This inferential analysis explores the `ToothGrowth` data set using the following steps:  

1. Basic exploratory data analysis
2. Summarize the data
3. Examine the effect of supp and dose on tooth growth using confidence intervals and hypothesis tests  
4. State conclusions and assumptions

###**Step 1: Exploratory data analysis**

```{r, message= FALSE, warning = FALSE, comment=""}
library(ggplot2)

data(ToothGrowth)

# Examine data set stucture
str(ToothGrowth)

# Determine number of observations
dim(ToothGrowth)

# Coerce dose variable to doses
ToothGrowth$dose <- as.factor(ToothGrowth$dose)

# Examine data structure after conversion
str(ToothGrowth)

```


###**Step 2: Summarize the data**
```{r,comment = ""}
# Summary statistics for all variables
summary(ToothGrowth)

# Table of dose levels and delivery methods
table(ToothGrowth$dose, ToothGrowth$supp)

# Exploratory plots
ggplot(ToothGrowth,aes(x = dose, y = len))+ geom_boxplot(aes(fill = dose))
ggplot(ToothGrowth,aes(x = supp, y = len)) + geom_boxplot(aes(fill = supp))
```

###**Step 3: Examine the effect of supp and dose on tooth growth using confidence intervals and hypothesis tests **

A T-test was performed to check for group differences explained by supplement type.
```{r,comment=""}
t.test(len ~ supp, data = ToothGrowth)
```
As shown above, the p value is 0.06 and the confidence interval contains 0. Thus, we can **not reject** the null hypothesis that different supplement types have no effect on tooth length.  

T-tests were then performed to determine if group differences are due to dose levels.

```{r,comment=""}

# Subgroup does by their level pairs
ToothGrowth_Dose_0.5_1.0<-subset(ToothGrowth, dose %in% c(0.5, 1.0))
ToothGrowth_Dose_0.5_2.0<-subset(ToothGrowth, dose %in% c(0.5, 2.0))
ToothGrowth_Dose_1.0_2.0<-subset(ToothGrowth, dose %in% c(1.0, 2.0))

# Group Differences due to 0.5 to 1.0 dose levels
t.test(len ~ dose, data = ToothGrowth_Dose_0.5_1.0)

# Group Differences due to 0.5 to 2.0 dose levels
t.test(len ~ dose, data = ToothGrowth_Dose_0.5_2.0)

# Group Differences due to 1.0 to 2.0 dose levels
t.test(len ~ dose, data = ToothGrowth_Dose_1.0_2.0)
```
For all three tests, the p value is less than 0.05 and the confidence intervals does not contain zero.  Thus, we can **reject** the null hypothesis and conclude that increasing dose levels ***does*** in fact lead to an increase in tooth length.  

###**Step 4: State conclusions and assumptions**
 
Based on my analysis above my conclusions are as follows:  
1. Supplement type does not effect tooth growth  
2. Increasing dose levels does lead to increased tooth growth.  

This conclusion is based on the following assumptions:  
- The experiment was conducted using random assignment to control for confounding factors   
- The sample population is representative of the entire population  
- Variance for the t-tests are different for the two groups being compared  
